---
title: 'Project 1: Exploratory Data Analysis'
author: "SDS348"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

## **Data Wrangling and Data Exploration**
**Broadway Data Analysis** 
*Project by: Ericka Salas (efs554)*  

    
## **Introduction of Broadway Data**

  Before the COVID-19 pandemic hit, I worked at a box office on UT Austin's campus that mainly sold tickets to concerts and broadway performances. I became a huge broadway fan and for the past two years of working in the office, attending broadway musicals has always been fun and exciting. I miss the feeling of the lights dimming at the beginning of a show or intermission conversations on the songs I enjoyed the most. Since I miss watching broadway musicals, I decided to analyze the different successes of different broadway shows from 19. 

  The data I gathered is from Github's "Tidy Tuesday" which specifically comes from Playbill, an offical broadway magazine and website. This data shows weekly box office grosses of revenue, attendance in different Broadway League theaters, weekly dates, average ticket price, potential gross, preview numbers, number of performances, etc. There is a total of fifteen numberic variable and two categorical variables. I am looking at two different datasets. One is composed of all the weekly grosses for each show while the other is composed of shows that started before 1985. I am mainly curious on what are the most successful broadway shows of all time.  Also, I want to analyze if ticket prices have changed since the premiere of each show. Overall, I expect popular broadway shows like *Hamilton, Cats, and The Lion King* to have the most revenue or performance attendance.

#### Loading necessary libraries
```{r}
library(tidyverse)
library(dplyr)
#importing this library to easily convert a date to a year by itself
library(lubridate)
```

#### Reading in data for weekly gross revenue, attendance, previews, etc.
```{r}
grosses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-28/grosses.csv', guess_max = 40000)
```

#### Reading in data for before 1985 shows
```{r}
pre_1985_starts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-28/pre-1985-starts.csv')
```

## **Tidying: Rearranging Wide/Long**

These datasets are already tidy, however, I untidyed it and  gathered it back together to demonstrate how the variables were spread out before. First I piped the data set to spread out the theater location variables with values from the week each performance stopped premiering. Subsequently, to tidy it back together, I gathered all the columns that were theater names and placed them in a column called theater along with its weekend final premier date. 

####Tidying the dataset
```{r}
grosses <- grosses %>% pivot_wider(names_from = theatre, values_from = week_ending) %>% pivot_longer(cols = "St. James Theatre":"Helen Hayes Theater", names_to="theatre", values_to ="week_ending")
head(grosses)
```

    
## Joining/Merging

I decided to do a left join since the common variable between the datasets is "show". The majority of the show data is in the grosses variable, which is the main focus of the data. Therefore, the left join omits the unneccessary variables that are present in the "pre_1985_starts" datasets and joins the shows in both. The cases that were dropped were the shows not present in the overall gross dataset. In other words, a few shows that premiered prior to 1985 and were not present in the gross dataset were dropped. 

#### Joining/Merging the Broadway Data
```{r}
broadway <- left_join(grosses, pre_1985_starts, by = "show")
```


##  **Wrangling**

#### Grouping the data by show and theater first then filtering by a certain popular broadway theater in New York City
```{r}
#selecting avg_ticket_price, show, theatre, then arranging the ticket prices by most expensive ticket price in theatre
highest_price <- broadway %>% group_by(show) %>% filter(theatre == "Majestic Theatre") %>% select(avg_ticket_price, show) %>% arrange(desc(avg_ticket_price))
head(highest_price)
```

#### Function to calculate the number of weeks since start and mutating the week of run orginial number using select() and mutate()
```{r}
calculate_weeks_since_start <- function(x) {
  as.integer(pmax(1, difftime("1985-06-09", x, units = "weeks")))
}

broadway <- broadway %>% mutate(week_of_run_originals = calculate_weeks_since_start(start_date) + row_number())
```

#### Counts the number of shows that have aired since it started and gets the top
```{r}
broadway %>% group_by(show) %>% summarize(count=n()) %>% arrange(desc(count)) %>% head
```

#### Calculates the mean number of weeks from the start of an original premiere date
```{r}
broadway %>% summarize(mean_original_week = mean(week_of_run_originals, na.rm=T)) 
```

#### This summarizes the top shows by year
```{r}
top_tickets <- grosses %>% mutate(year = year(week_ending)) %>% filter(year >= 1996, !is.na(top_ticket_price)) %>% group_by(year, show) %>% summarise(avg_ticket_price = mean(avg_ticket_price, na.rm = TRUE), avg_top_ticket_price = mean(top_ticket_price, na.rm = TRUE)) %>% arrange(desc(avg_top_ticket_price))
top_tickets %>% head
```

#### Getting the shows with the most seats 
```{r}
broadway %>% group_by(show) %>% summarize(mean_show_seats_sold = mean(seats_sold
,na.rm=T), sd_show_seats_sold = sd(seats_sold, na.rm=T)) %>% arrange(desc(mean_show_seats_sold)) %>% head
```

#### Getting the theatre that sold the most seats
```{r}
broadway %>% group_by(theatre) %>% summarize(mean_theatre_seats_sold=mean(seats_sold
,na.rm=T), sd_theatre_seats_sold =sd(seats_sold, na.rm=T)) %>% head
```

#### Getting the median, max, and min average ticket price for the show Cats
#### These are three out of the five unique functions needed to be used in summarize 
```{r}
broadway %>% filter(show == "Cats") %>% summarize(med_cats_price = median(avg_ticket_price, na.rm=T), min_cats_price = min(avg_ticket_price, na.rm=T), max_cats_price =max(avg_ticket_price, na.rm=T))
```

### Summary of Results
After summarizing different variables and grouping them by categoricals, the results of the Broadway shows and theatres appeared to be interesting. The Broadway performance with the highest average ticket price of all time in the Majestic Theatre is Springsteen on Broadway with a price of 511.58 US dollars. *The Phantom of the Opera* appeared to be the show with the most number of performances with a total of 97266 premieres. The mean number of weeks that have passed since a show has premiered is 582716.9, which means that many of the shows in this data are successful and continue to premiere. *Hamilton* appeared to be the top show for five years straight, from 2016 to 2020. In addition, the Broadway show that had the most attendance and seats sold ever in a single theater was *David Copperfield: Dreams and Nightmares*, which had a value of 20043 seat sold. On average, all the Broadway League theatre sold approximately 7893 seats per show. 

Since *Cats* is one of the most classic and popular Broadway shows, I decided to retrieve the median, minimum, and max average price of tickets. The median ticket price is $42.89 and the minimum is 35.59. On average, the maximum ticket price of *Cats* is 147.17, which means that this performance is less costly than other shows such as *Hamilton*. 

## **Visualizing**

### Correlation HeatMap of Numeric Variables
```{r}
#Getting only the numeric values 
cormat <- broadway %>% na.omit() %>% select_if(is.numeric) %>% cor(use="pair")

#Making the correlation matrix tidy
tidycor <- cormat %>% as.data.frame %>% rownames_to_column("var1") %>% pivot_longer(-1,names_to="var2",values_to="correlation")

#Creating the correlation heatmap
tidycor %>% ggplot(aes(var1,var2,fill=correlation))+
geom_tile() + scale_fill_gradient2(low="cyan",mid="white",high="red") +  geom_text(aes(label=round(correlation,2)),color = "black", size = 2) + theme(axis.text.x = element_text(angle = 90, hjust=1))+coord_fixed() + labs(title = "Correlation HeatMap")
```

### Plot Number One - Top Ticket Prices for the Past 24 years

```{r}
#mapping the three variables
broadway %>% ggplot(aes(year, avg_ticket_price, color = theatre))  + geom_line(data = top_tickets, size = 1.5, col = "blue", stat = "Summary") + geom_point(data = top_tickets, size = 1.5, col = "white", stat = "Summary") + theme_dark() +
  #fixing the labels here (x, y, title)
  scale_x_continuous("Year", breaks = seq(1984, 2020, 2)) + scale_y_continuous("Average ticket Price (USD $)") + labs(title = "In the past 24 years, the cost of a Broadway tickets doubled.", subtitle = "Inflation-adjusted average ticket prices (Jan. 2020 dollars)", caption = paste("Source: Playbill", "Points are average ticket price per show",sep = "\n"), x = "", y = "") + 
  #adding a new theme outside the ones incorporated
  theme(text = element_text(family = "Bahnschrift"), panel.grid.minor = element_blank())
```

###Plot Number Two - Top Ticket Sales by Price in 2020
```{r}
#subsetting a dataframe to be the top tickets of only 2020
top_2020 <- top_tickets %>% filter(year == 2020)
  #mapping three diff variables
top_2020 %>% ggplot(aes(x = show, y = avg_top_ticket_price, color = as.factor(theatre))) +
  #adding a theme and setting the x, y, and title labs
  theme_light() + geom_bar(data = top_2020, size = 1.5, col = "dark blue", stat = "Summary") + labs(title = "The Average Top Ticket Prices of the Top Broadway Shows in 2020" ,subtitle = "Inflation-adjusted average ticket prices (Jan. 2020 dollars)", caption = paste("Source: Playbill")) + ylab("Average Top Ticket Price ($)") + xlab("Broadway Show") + theme(axis.text=element_text(size=6.5), axis.title=element_text(size=12,face="bold")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, face = "bold"))
```
The first plot shows that the average ticket price per broadway show has increased in the past 24 years. In 1996, the average price per ticket was less than 50 dollars. The price increased per year, conveying a  positive correlation between year and average ticket price. In 2020, the average ticket price increased to above 100 US dollars. There are a few shows that bring the 2020 prices higher since they are in demand such as Hamilton, Dear Evan Hansen, etc.

The second plot shows the top ticket prices of 2020 before the COVID-19 pandemic. Therefore, this data is not a full year, but the 2020 season from December to late March. This is a reason why A Christmas Carol made onto this plot since it was mainly performed during the holiday season. The show with the highest ticket price is Hamilton, which approximately reaches as high as 1,000 USD. Waitress appears to be the second highest with a max of around 600 USD. The rest of the shows appear to be the same due to the uniform distribution of the plot. The majority of shows had a high price that reach up to 400 USD. This plot shows that these broadway performance were the most paid for and had the most purchased tickets with the above price ranges.

    
## **Dimensionality Reduction**

The chosen numeric variables

####Taking the dataset and preparing it
```{r}
pre_1985_starts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-28/pre-1985-starts.csv')
grosses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-04-28/grosses.csv', guess_max = 40000)
broadway <- left_join(grosses, pre_1985_starts, by = "show")
broadway <- broadway %>% na.omit()
```

#### Creating correlation matrix and performing PCA
```{r}
broadway1 <- broadway %>% select(-show,-theatre)
nums <- broadway1 %>% select_if(is.numeric) %>% scale
rownames(nums)<- broadway1$show
broad_pca <- princomp(nums)

datScale <- broadway %>% select(is.numeric) %>% scale
eig1 <- broadway %>% select(is.numeric) %>% cor() %>% eigen()
PCAscores <- datScale %*% eig1$vectors
```

####Choose PCA numbers to keep
```{r}
eigval <- broad_pca$sdev^2 #square to convert SDs to eigenvalues
varprop=round(eigval/sum(eigval), 2) #proportion of var explained by each PC
```

####Grab PC Scores
```{r}
#using cumulative property of variance
round(cumsum(eigval)/sum(eigval), 2)
```
Therefore, 2 or 3 PC's are most ideal

####Visualize and interpret
```{r}
broadway %>% mutate(PC1 = PCAscores[,1], PC2 = PCAscores[,2], PC3 = PCAscores[,3])%>%
ggplot(aes(PC1,PC2,color=theatre))+geom_point()
```

Since we want to know what direction the data varies the most, a Principle Component Analysis. First a correlation matrix was created by scaling the data and retrieving only the numeric variables. Three PC numbers were kept after doing a cumulative proportion of variance. Based on the PCA score plot above, the two theatres that are most clearly separate by PC2 are *Ford Center for the Performing Arts* and *Longacre Theater*. In additon, *Vivian Beaumont Theater*, *Gerald Scholenfeld Theatre*, and *Ford Center for Performing Arts* are clearly separate by PC1. There are a few outliers in the clusters, but they are mainly from the *Hudson Theatre*.



